---
title: "p8105_hw2_mk5149"
author: "Mungyu Kwok"
date: "2025-09-30"
output: github_document
---
# Preparation
Firstly, we will load any necessary packages.
```{r}
library(tidyverse)
library(lubridate)
```

Next, we will load the datasets (Fivethirtyeight), and save them into dataframe.
```{r}
pols_df = read_csv("fivethirtyeight_datasets/pols-month.csv")
snp_df = read_csv("fivethirtyeight_datasets/snp.csv")
unemployment_df = read_csv("fivethirtyeight_datasets/unemployment.csv")
```

# Problem 1
First, clean the data in `pols-month.csv`. 
We will use the `separate()` function to break up the variable `mon` into integer variables `year`, `month`, and `day`.
Then, we will replace month number with month name; 
We will also create a president variable taking values `gop` and `dem`, and remove `prez_dem` and `prez_gop`; 
Finally, we will remove the `day` variable.
```{r}
pols_clean = 
  pols_df |>
  separate(mon, into = c("year","month","day"), sep = '-', convert = TRUE) |> 
  mutate(
    month = month.abb[month], #buildt-in R constant
    president_party = case_when(
      prez_gop == 1 ~ "gop",
      prez_dem == 1 ~ "dem"
    )
  ) |>
  select(-day, -prez_gop, -prez_dem) # Removes only the day, prez_gop, and prez_dem columns

```

Secondly, we will clean the data in `snp.csv` by using a similar process to the above. 
For consistency across datasets, arrange according to year and month, and organize so that `year` and `month` are the leading columns.
```{r}
snp_clean = 
  snp_df |>
  mutate(
    date = mdy(date),
    year = year(date),
    month = month(date, label = TRUE, abbr = TRUE) |>
      as.character()
  ) |>
  select(year, month, snp_close = close)
```

Thirdly, we will tidy the unemployment data so that it can be merged with the previous datasets. 
This process will involve switching from “wide” to “long” format; 
ensuring that key variables have the same name; 
and ensuring that key variables take the same values.
```{r}
unemployment_clean = 
  unemployment_df |>
  pivot_longer(
    cols = Jan:Dec,
    names_to = "month",
    values_to = "unemployment_rate"
  ) |> 
  rename(year = Year) # Add this line to fix the capitalization, which will cause error in the later part (merging the data)
```

For the final part of problem 1, we will join the datasets by merging `snp` into `pols`, and merging `unemployment` into the result.
```{r}
fivethirtyeight_df = 
  pols_clean |>
  left_join(snp_clean, by = c("year","month")) |>
  left_join(unemployment_clean, by = c("year", "month"))
#left_join: based on shared columns, if no matches -> NA

# View a glimpse of the final merged data
glimpse(fivethirtyeight_df)
```










