---
title: "p8105_hw2_mk5149"
author: "Mungyu Kwok"
date: "2025-09-30"
output: github_document
---
# Preparation
Firstly, we will load any necessary packages.
```{r}
library(tidyverse)
library(lubridate)
```

Next, we will load the datasets (Fivethirtyeight), and save them into dataframe.
```{r}
pols_df = read_csv("fivethirtyeight_datasets/pols-month.csv")
snp_df = read_csv("fivethirtyeight_datasets/snp.csv")
unemployment_df = read_csv("fivethirtyeight_datasets/unemployment.csv")
```

# Problem 1
First, clean the data in `pols-month.csv`. 
We will use the `separate()` function to break up the variable `mon` into integer variables `year`, `month`, and `day`.
Then, we will replace month number with month name; 
We will also create a president variable taking values `gop` and `dem`, and remove `prez_dem` and `prez_gop`; 
Finally, we will remove the `day` variable.
```{r}
pols_clean = 
  pols_df |>
  separate(mon, into = c("year","month","day"), sep = '-', convert = TRUE) |> 
  mutate(
    month = month.abb[month], #buildt-in R constant
    president_party = case_when(
      prez_gop == 1 ~ "gop",
      prez_dem == 1 ~ "dem"
    )
  ) |>
  select(-day, -prez_gop, -prez_dem) # Removes only the day, prez_gop, and prez_dem columns

```












